---
title: "Ex3DA2223_Alzbeta_Reinhardova"
author: "Alzbeta_Reinhardova_6180079"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1.1.1 General questions on moderation 
##1.
moderation = interaction -> regression analysis and is a relationship between two variables when they depend on another third variable (moderator variable)

##2.
main effect is between the independent and dependet variable -> e.g. time of day (day, night), number of phone calls, gender (moderator, man, woman, non-binary)
main effect = would be comparison whether there is more phone calls during the day or night
simple effect = would be which gender call more during day and night 
main effect = first order effect 

##3.
to avoid multicollinearity issues that may disturb our outcomes and make it not valid
to have better interpretation of the results 

##4.
the increase of a grade is 0.018 for every consumed coffee -> that is what the beta means in this contex

##5.
anova and regression are both linear models => the difference is that regression is a method that is looking for a relationships between variables to make predictions of the DV (dependent) based on IV (independent) 
ANOVA is a statistical tool applied to unrelated groups looking for a commmon mean = analysis of variance 
only for categorical = anova
also for continuous = ancova 

##6.

Beta 1 and beta 2 are the lower (use instead of lower he likes it  = first) order effects of X1 and X2 while the interaction effect beta 3 is a higher order effect.


#1.1.2 Interaction with categorical predictors 
##1.
Factorial designs -> often used Anova 

##2.
no -> control is a group and is not a factor = some experiments = within-subject -> has control group within so I look at before treatment and after treatment  instead of companring outcomes between groups 

##3. 
lm(outcome ~ p1 * p2, data = data)

##4.
```{r}
install.packages("tidyverse")
df_news <- read.csv(file = "news.csv", header=TRUE, sep = ",")
df_news
df_news_filtered <- filter(df_news, section%in%c("sport", "news")) %>% select(-id)

```

##5.
name of design = factorial design
```{r}
section_factor <- as.factor(df_news_filtered$section)
day_factor <- as.factor(df_news_filtered$day)

#day_f <- c("monday", "friday")
#d_factor <- factor(day_f)

table(section_factor)
table(day_factor)

#levels(df_news_filtered$day) = only if I have leves


WHY IS IT NOT WORKING? 
#outcome_y <- lm(y ~ day * news, data = df_news)
#summary(df_news_filtered)


#df_news$day <- factor(df_news$day, levels = c(0,1), labels = c("Monday", "Friday"))
#df_news$section <- factor(df_news$section, levels = c(0,1,2), labels = ("news", "sport","business"))
https://stackoverflow.com/questions/9251326/convert-data-frame-column-format-from-character-to-factor 
```

##6.
```{r}
#outcome_y <- lm(y ~ section, df_news)
#outcome_y
outcome_2 <- lm(y ~ section + day + section:day, data = df_news)
outcome_2
# should have used the factored data set ! -> NEED summary(outcome_2)
```
INTERPRETATION = so if we actually look below we can see that people watch more news on Friday than on Monday -> busienss is most popular and news are least popular
are these findings significant? = 
the main effect is between days -> Monday and Friday amount of sections watched 
the simple effect is between news, business, and sport
**IF YOU WANT INSTEAD OF SPORT SEEING NEWS-> CHANGE IT TO THE 0 POSITION -> 0 = NEWS AND 1 = SPORTS => LOOKING IN NEWS -> 0= SPORTS AND 1= NEWS -> LOOKING AT SPORTS**

##7.


```{r}
install.packages("interactions")
interactions::cat_plot(outcome_2, pred = section, modx = day, plot.points = T)
```


##8.
```{r}
install.packages("emmeans")
comp <- emmeans(outcome_2, ~ section * day)
comp
#contrast (jlksj , ''bon'')



```

#1.1.3 Interaction with continous and categorical predictors
##1.

```{r}
df_distraction <- read.csv(file = "distraction.csv", header=TRUE, sep = ",")
df_distraction
summary(df_distraction)
```
##2.
```{r}
install.packages("ggplot2")
df_distraction_filtered <- filter(df_distraction, group%in%c("urban"))
df_distraction_filtered

plot(df_distraction_filtered$dist_level, df_distraction_filtered$simple_rt, main = "Relationship", xlab = "simple_rt", ylab = "dist_level")
#abline(lm(dist_level ~ simple_rt, data = df_distraction_filtered), col= "blue")

GEOM_SMOOTR()???
```
INTERPRETATION = from the graph we can see that the higher the distraction level the higher the reaction time -> the relationship seems to be linear -> that is what the graph indicates however if we want to be sure we need to run more tests

##3.
```{r}
model_one <- lm(simple_rt ~ dist_level+group, data = df_distraction)
model_one
summary(model_one)
```
INTERPRETATION = 

##4.
```{r}
#reaction_centered <- scale(df_distraction$simple_rt)
distraction_centered <- scale(df_distraction$dist_level)
model_two <- lm(simple_rt ~ distraction_centered * group, data = df_distraction)
summary(model_two)
```
INTERPRETATION =  

##5.
```{r}
interactions::cat_plot(model_two, pred = distraction_centered, modx = group, plot.points = T )
interactions::interact_plot(model_two, pred = distraction_centered, modx = group, plot.points = TRUE)

```
INTERPRETATION = 

#1.1.4 AN(C)OVA environments
IN THE PREVIOUS EXAM

##1.

```{r}
df_brew <- read.csv(file = "brew.csv", header=TRUE, sep = ",")
new_df_brew <- df_brew[-c(1),]
new_df_brew

#new_df_brew$taste <- as.numeric(as.character(new_df_brew$taste))
class(new_df_brew$taste)

df_brew_filtered <- filter(df_brew, attention%in%c("1"))

sapply(new_df_brew, class)

summary(df_brew_filtered)
#ggplot(brew, aes(x=brew, y=taste)) + geom_point(shape = 18, colour = "blue") + geom_smooth(method=lm, se = FALSE)

#YOU HAD TO MAKE A GRAPH AFTER BEFORE REMOVING 1 ROW 
```
INTERPRETATION = 

##2.
independent = sample = between subject 
independent = brew = within (if the customer taste original and after local brew)
within => repeated measure => if the customer would try first original and later on local brew = the brew would be repeated measure -> within subject 

##3.

```{r}
outcome_anova <- lm(taste ~ sample + brew, data = new_df_brew)
outcome_anova

install.packages("afex")
outcome_anova_two <- aov_ez("sample", "taste", new_df_brew, between = "sample", within =  c("original", "local"))

comp <- emmeans(outcome_anova_two, ~ sample * brew)
comp

```


##4.

```{r}

afex_plot(outcome_anova_two, "sample", "original", "local")
```
https://cran.r-project.org/web/packages/afex/readme/README.html

https://tysonbarrett.com/jekyll/update/2018/03/14/afex_anova/